Supplementary Information

### 2021-03-23 doc/manuscript/ForestGEO_treerings.Rmd
EGA feedback

*(keep this paragraph mostly as is):*
Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it has changed through time [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020].
Combined with forest censuses, they can be used to estimate forest woody productivity, $ANPP_{woody}$ [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
Tree rings also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.

@@

*(reword to ecology focus):*
Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
Over timeframes of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad climatic gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided (e.g., by sampling at sites with characteristics that result in trees being more climate-limited), and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon. Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in tree growth responses to interannual variation in climate that occur for trees within forest settings.
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**

@@

All tree cores **(n = XXXX)** were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
From among the full set of original RW measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis. 
We also excluded records with small sample size or highly anomalous growth patterns, including
(1) species with < 7 cores,
(2) cores with < 30 years of record,
(3) contiguous portions of cores containing large outliers (RW > mean plus 5 x SD of RW for the entire core),
and (4) the final 20 years prior to death of trees cored dead. 
The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
From analyses including DBH (see below), we further excluded
(1) trees for which we lacked data required to reconstruct DBH,
(2) trees for which there was a significant inconsistency between measured DBH and the sum of RW's across the core (Appendix `r DBH_reconstruction`),
and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) included < 3 conspecific trees.
In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with DBH (Table `r sampling_details_table`).

For each year in the tree-ring records, we reconstructed DBH, as detailed in Appendix `r DBH_reconstruction`. 
We used allometric equations between DBH and bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).

Once DBH had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass (AGB). 
Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$].

Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on more local records (Appendix `r climate_correction`). 
Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation (PPT); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration (PET) and precipitation day frequency (PDF).
For the one riparian site, NIO, we tested for an effect of stream flow (SF), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S. Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw). 
All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 [@anderson-teixeira_forestgeo_2020].

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, thereby producing residuals. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
We then used *climwin* to identify the climate variables that most strongly correlated to the residuals of RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): *residual growth index ~ [climate] + (1 | sp) + (1 | treeID)*.
Here, for each permutation, *climate* specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables, we selected the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.

@@

Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and DBH (models with DBH and its climate interactions) into linear mixed-effects models (function *lme* in R package *nlme*), with core identity as a random intercept and yeaar as a continuous time covariate for the within-group correlation structure (function *corCAR1*). We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013]. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta$AIC for each. The time window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model). Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)

@@

![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** *Climwin*-selected climate variables are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential. For each species (color-coded as in Fig. 4), relationships are plotted if included in top model. For each relationship shown, other terms in the model are held constant at their median. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)

@@

###other:
n_pos_first_order_P_term=42 #(exception: TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
```
Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species, and were predominantly positive (Fig. `r clim_sens`).
PPT was selected over PDF as the top variable in five of the eight sites with both variables available (but had no significant main effect at one of these, NIO), and was the only option at two sites (LT and CB).
Optimal time windows most commonly coincided at least partially with months of active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer in the extratropics (n= 5 of 8).
In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
In the extratropics, optimal windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to PPT or PDF were predominantly positive, with positive first-order linear terms for all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig. `r clim_sens`).

@@

All three growth metrics, RW, BAI, and $\Delta$AGB, varied with DBH for most species at all sites (Fig. `r size`).
For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear DBH term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative). 
There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig. `r size`).
At the other end of the spectrum, some species had low RW at small DBH, increased to peak RW at intermediate DBH, and subsequently declined.
These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp. at SCBI and Žofín, *Acer saccharum* at LDW, *Picea* spp. at Žofín and CB; Table `r species_table`).

@@

Effects of year and DBH interacted such that inclusion of year in in models altered the shape of DBH responses, typically resulting in less pronounced growth declines with increasing DBH (Figs. `r all_GLS`).

@@

The tree-ring analysis approach developed here allowed simultaneous consideration of the effects of primary climate drivers, tree size, and slowly changing environmental conditions on annual growth across multiple forest types and species (Fig. `r schematic`), yielding results that are consistent with those that would be obtained using established methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a more complete picture of how climate, tree size, and year have shaped tree growth at 10 globally distributed forest over the past 120 years.

@@

**(should I remove this paragraph?)**
While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
First, in determining the primary climate drivers (step 1; Fig. `r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged. 
Some differences in variables identified and the slope between growth and climate are to be expected given the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`). 
~~Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the primary climate drivers (step 1; Fig. `r schematic`).~~ 
~~This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf. Figs. `r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually be species.~~ *(2nd sentence moved to methods)*

Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were prevalent Fig. `r clim_sens`). 
This result, which is consistent with physiological expectations [REFS, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear. 
The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], are likely to become more prevalent [@germain_climate_2020].

@@

We found evidence of saturation or decline in BAI and $\Delta$AGB for the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed (Fig. `r size`), contrasting with findings of cross-sectional analyses showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree rings [@foster_predicting_2016].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r all_GLS`), suggesting that some of the declines in Figure `r size` are more properly attributed to the effect of year than DBH.

>>>>

*(keep this paragraph mostly as is):*
Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it has changed through time [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020; @au_demographic_2020].
Combined with forest censuses, they can be used to estimate forest woody productivity, $ANPP_{woody}$ [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
Tree rings also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.

@@

*(reword to ecology focus):*
Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
Over time frames of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad climatic gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided (e.g., by sampling at sites with characteristics that result in trees being more climate-limited), and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
*Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon. Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in tree growth responses to interannual variation in climate that occur for trees within forest settings.*
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**

@@

All tree cores **(n = XXXX)** were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
From among the full set of original RW measurements, we excluded cores for which we detected technical errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis. 
We also excluded records with small sample size or highly anomalous growth patterns, including
(1) species with < 7 cores,
(2) cores with < 30 years of record,
(3) contiguous portions of cores containing large outliers (RW > mean plus 5 x SD of RW for the entire core),
and (4) the final 20 years prior to death for trees cored dead. 
The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
From analyses including DBH (see below), we further excluded
(1) trees for which we lacked data required to reconstruct DBH,
(2) trees for which there was a significant inconsistency between measured DBH and the sum of RW's across the core (Appendix `r DBH_reconstruction`),
and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) included < 3 conspecific trees.
In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with DBH (Table `r sampling_details_table`).

For each year in the tree-ring records, we reconstructed DBH, as detailed in Appendix `r DBH_reconstruction`. 
We applied allometric equations for bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).

Once DBH had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass (AGB). 
Biomass allometries for temperate and tropical species were calculated using the R packages *allodb* (Gonzalez-Akre et al. in revision) and *BIOMASS* [@rejou-mechain_biomass_2017], respectively.
We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$].

Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on higher-resolution or local records (Appendix `r climate_correction`). 
Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation (PPT); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration (PET) and precipitation day frequency (PDF).
For the one riparian site, NIO, we tested for an effect of stream flow (SF), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S. Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw). 
All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 [@anderson-teixeira_forestgeo_2020].

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, thereby producing residuals. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
We then used *climwin* to identify the climate variables that most strongly correlated to the residuals of RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): *residual growth index ~ [climate] + (1 | sp) + (1 | treeID)*.
Here, for each permutation, *climate* specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables, we selected the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.

@@

Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and DBH (models with DBH and its climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and yeaar as a continuous time covariate for the within-group correlation structure (function *corCAR1*). We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013]. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta$AIC for each. The time window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows the GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model). Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)

@@

![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential. For each species (color-coded as in Fig. 4), relationships are plotted if included in top model. For each relationship shown, other terms in the model are held constant at their median. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)

@@

###other:
n_pos_first_order_P_term=42 #(exception: TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
```
Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species, and were predominantly positive (Fig. `r clim_sens`).
PPT was selected over PDF as the top variable in five of the eight sites with both variables available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
Optimal time windows most commonly coincided at least partially with months of active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer in the extratropics (n= 5 of 8).
In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
In the extratropics, optimal windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to PPT or PDF were predominantly positive, with positive first-order linear terms for all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig. `r clim_sens`).

@@

All three growth metrics, RW, BAI, and $\Delta$AGB, varied with DBH for most species at all sites (Fig. `r size`).
For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear DBH term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative). 
There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig. `r size`).
At the other end of the spectrum, some species had low RW at small DBH, increased to peak RW at intermediate DBH, and subsequently declined.
These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp. at SCBI and ZOF, *Acer saccharum* at LDW, *Picea* spp. at ZOF and CB; Table `r species_table`).

@@

Effects of year and DBH interacted such that inclusion of year in models altered the shape of DBH responses, typically resulting in less pronounced growth declines with increasing DBH (Figs. `r all_GLS`).

@@

The tree-ring analysis approach developed here allowed simultaneous consideration of the effects of primary climate drivers, tree size, and slowly changing environmental conditions on annual tree growth across multiple forest types and species (Fig. `r schematic`), yielding results that are consistent with those that would be obtained using established methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) but offering a more complete picture of how climate, tree size, and year have shaped tree growth at 10 globally distributed forest over the past 120 years.

@@

**(should I remove this paragraph?)**
While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
First, in determining the primary climate drivers (step 1; Fig. `r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged. 
Some differences in variables identified and the slope between growth and climate are to be expected given the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`). 
~~Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the primary climate drivers (step 1; Fig. `r schematic`).~~ 
~~This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf. Figs. `r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.~~ *(2nd sentence moved to methods)*

Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were prevalent (Fig. `r clim_sens`). 
This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear. 
The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], are likely to become more prevalent [@germain_climate_2020].

@@

We found evidence of saturation or decline in BAI and $\Delta$AGB for the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed (Fig. `r size`), contrasting with findings of cross-sectional analyses showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree rings [@foster_predicting_2016].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r all_GLS`), suggesting that some of the declines (Fig. `r size`) are more properly attributed to the effect of year than DBH.
